---
title: "Apply ctSVG to Visium HD data"
author: "Haotian Zhuang"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Apply ctSVG to Visium HD data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
date: '`r format(Sys.Date(), "%B %d, %Y")`'
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction
We will demonstrate how to use ctSVG to generate single-cell gene expression profiles and identify cell-type-specific SVGs from Visium HD data. The Visium HD dataset of a mouse brain tissue was downloaded from the [10x website](https://www.10xgenomics.com/datasets/visium-hd-cytassist-gene-expression-libraries-of-mouse-brain-he).

# Download Space Ranger output files and H\&E image
The Space Ranger output files were downloaded at <https://cf.10xgenomics.com/samples/spatial-exp/3.0.0/Visium_HD_Mouse_Brain/Visium_HD_Mouse_Brain_binned_outputs.tar.gz>. The H\&E image was downloaded at <https://cf.10xgenomics.com/samples/spatial-exp/3.0.0/Visium_HD_Mouse_Brain/Visium_HD_Mouse_Brain_tissue_image.tif>.

# Perform nuclei segmentation
The nuclei segmentation results can be obtained by running segmentation methods, such as [**StarDist**](https://github.com/stardist/stardist), on the H\&E image accompanying the Visium HD data.

```
import pandas as pd
import numpy as np
from tifffile import imread, imwrite
from csbdeep.utils import normalize
from stardist.models import StarDist2D

img = imread('Visium_HD_Mouse_Brain_tissue_image.tif')

# Load the pretrained model
model = StarDist2D.from_pretrained('2D_versatile_he')

# Percentile normalization of the image
min_percentile = 5
max_percentile = 95
img = normalize(img, min_percentile, max_percentile)

# Stardist
labels, polys = model.predict_instances_big(img, axes='YXC', block_size=4096, prob_thresh=0.01,nms_thresh=0.001, min_overlap=128, context=128, normalizer=None, n_tiles=(4,4,1))

# Convert data format
data = [(cell_id, x, y) for cell_id, (xs, ys) in enumerate(polys['coord']) for x, y in zip(xs, ys)]
df = pd.DataFrame(data, columns=['Cell', 'X', 'Y'])

df.to_csv('nuclei.csv', index=False)
```

# Load package
```{r setup}
library(ctSVG)
mydir <- "~/Desktop/Research/ctSVG/mouse_brain"
list.files(mydir)
```

# Assign each square to a cell
```{r}
suppressMessages(suppressWarnings(
  cellassignRes <- cellassign(visiumpath = file.path(mydir, "binned_outputs/square_002um"),
                              segmentationpath = file.path(mydir, "nuclei.csv"))
))
```

```{r eval=FALSE, include=FALSE}
cellassignRes_ref <- readRDS(system.file("extdata", "cellassignRes.rds", package = "ctSVG"))
all.equal(cellassignRes, cellassignRes_ref)
```

```{r}
str(cellassignRes)
coord <- cellassignRes$coord
```

# Generate single-cell gene expression profiles
```{r}
count <- getexpr(cellassignRes = cellassignRes,
                 visiumpath = file.path(mydir, "binned_outputs/square_002um"))
str(count)
```

# Run Seurat pipeline
```{r}
suppressMessages(suppressWarnings(
  d <- runseurat(count = count)
))
```

# Reassign cells to clusters
In this vignette, `n.max` is set to 200 to reduce computational time. By default, `n.max` is set to 1,000 in practice.

```{r}
recluRes <- recluster(d = d, n.max = 200)
```

# Identify spatially variable genes within each cell cluster
```{r}
suppressMessages(suppressWarnings(
  clutestRes <- ctsvg_test(d = d, coord = coord, recluRes = recluRes)
))
head(clutestRes)
```

# Fit spatial gene expression patterns within each cell cluster
```{r}
clufitRes <- ctsvg_fit(d = d, coord = coord)
```

# Visualize spatial patterns of a single gene
```{r}
plotGene(d = d, clufitRes = clufitRes, coord = coord, gene = clutestRes$gene[1],
         cellclu = "0", background = TRUE)
```

# Identify gene modules within each cell cluster
```{r}
moduleRes <- ctsvg_module(d = d, clutestRes = clutestRes, clufitRes = clufitRes)
```

# Visualize spatial patterns of metagenes within a cell cluster
```{r}
plotMetagene(d = d, clufitRes = clufitRes, moduleRes = moduleRes, coord = coord,
             cellclu = "0", background = TRUE)
```

<details>
  <summary>**Session Info**</summary>
```{r}
sessionInfo()
```
</details>
